---
title: "Advanced Data Visualization"
author: "Sami Gibson"
date: "2026-02-27"
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

**Set up Chunk**
```{r}
#| label: Set Up Chunk
#| message: false
#| warning: false

library(tidyverse) # loads in tidyverse package
library(janitor) # loads in janitor package
library(lubridate) # loads in lubridate package
library(here) # loads in here package
library(scales) # loads in scales package
library(showtext) # loads in showtext package
library(sysfonts) # loads in sysfonts package
library(ggtext) # loads in ggtext package
library(ggplot2) # loads in ggplot2 package
library(viridis) # loads in viridis package
library(fontawesome) # loads in fontawesome package

# Stores my personal data as an object called my_data
raw_data <- read.csv(here('data',"193DS data - stress (5).csv")) 
```
**Cleaning data**
```{r}
# Create a cleaned version of the raw dataset
cleaned_data <- raw_data |> 
  # Standardize column names:converts all column names to lowercase with underscores
  clean_names() |> 
  mutate(
    # Convert date column from character (mm/dd/yyyy) to proper Date format
    date = mdy(date_mm_dd_yyyy),
    # Convert day_of_week into an ordered factor
    day_of_week = factor(day_of_week, 
                         levels = c("Monday",
                                    "Tuesday", 
                                    "Wednesday",
                                    "Thursday",
                                    "Friday",
                                    "Saturday",
                                    "Sunday")),
    # Convert sleep efficiency from % style character to numeric proportion 
    sleep_efficiency = parse_number(as.character(sleep_efficiency)) / 100,
    # Convert Pilates variable to categorical factor
    pilates = factor(pilates, levels = c("No","Yes")))
```

Stores important info that changes as data set updates:

```{r}
# Calculate the total number of observations (days recorded)
n_obs <- nrow(cleaned_data)
# Identify the earliest date in the dataset
min_date <- min(cleaned_data$date, na.rm = TRUE)
# Identify the most recent date in the dataset
max_date <- max(cleaned_data$date, na.rm = TRUE)
```

Create base theme + aesthetics (xxx might change later)
```{r}
# Loads custom fonts 
# "Playfair Display" will be used for titles
# "Space Grotesk" will be used for axis labels and body text 
font_add_google("Playfair Display", "playfair")
font_add_google("Space Grotesk", "space")
# Activate showtext so ggplot can render fonts
showtext_auto()

# Creates custom gradient palette (used for stress, bins, heatmaps)
palette_gradient <- c(
  "#A6CEE3",  # light blue
  "#1F78B4",  # medium blue
  "#08306B"   # deep blue
)
palette_gradientA <- c(
  "#F7F7F7",  # very light
  "#A6CEE3",  # light blue
  "#1F78B4",  # medium blue
  "#08306B"   # deep blue
)
# Create a reusable base theme for all plots
base_theme <- theme_minimal(base_family = "space") +
  theme(
    # Main title styling 
    plot.title = (element_text(face = "bold", size = 18)),
    # Subtitle styling
    plot.subtitle = element_text(size = 11, margin = margin(b = 8)),
    # Caption styling
    plot.caption = element_text(size = 9, color = "gray30"),
    # Axis title styling
    axis.title = element_text(size = 11),
    # Axis tick label styling
    axis.text = element_text(size = 10, color = "gray20"),
    # Legend title styling
    legend.title = element_text(size = 10),
    # Legend text styling
    legend.text = element_text(size = 9),
    # Legend position
    legend.position = "bottom",
    # Legend direction
    legend.direction = "horizontal",
    # Removes minor grid lines
    panel.grid.minor = element_blank(),
    # Keeps subtle major grid lines
    panel.grid.major = element_line(linewidth = 0.25, color = "gray92"),
    # Adds margin around the edges of the plot
    plot.margin = margin(10, 12, 10, 12))
```

# Plot 1: Boxplot + jitter (Sleep by day of week)

```{r}
# Compute overall median sleep duration across all days (for reference line)
overall_median_sleep <- median(
  cleaned_data$time_slept_minutes, # numeric vector of sleep minutes
  na.rm = TRUE) # ignore missing values

# Define a custom color palette for each day of the week
week_colors <- c(
  "Monday" = "darkgoldenrod1",
  "Tuesday" = "mediumvioletred",
  "Wednesday" = "darkolivegreen",
  "Thursday" = "cyan4",
  "Friday" = "indianred",
  "Saturday" = "darkorange2",
  "Sunday" = "lightblue")

# Compute median sleep for each day of the week
med_by_day <- cleaned_data |> # uses cleaned data
  group_by(day_of_week) |> # group observations by weekday
  summarize(med_sleep = median(time_slept_minutes, # compute median sleep within each day
                               na.rm = TRUE), # remove missing values if any
            .groups = "drop") # ungroups df

# Identify the day with the highest median sleep
highlight_day <- med_by_day |>
  slice_max(med_sleep, n = 1) |> # select row with maximum median sleep
  pull(day_of_week) |> # extract just the weekday value
  as.character() # convert to character

# Store the numeric value of the highest median for annotation text
highlight_median <- med_by_day |>
  filter(day_of_week == highlight_day) |> # selects row with the highest median
  pull(med_sleep) # returns the med_sleep column as a numeric vector


# Build the boxplot visualization
p1_box <- 
  ggplot(
    cleaned_data, # uses cleaned data
    aes(x = day_of_week, # x axis is categorical predictor (weekday)
        y = time_slept_minutes)) + # y axis is response variable (sleep duration)
  geom_hline( # Add horizontal line at overall median sleep
    yintercept = overall_median_sleep, # location of reference line at median
    linetype = "dashed", # styles line as dashed
    linewidth = 0.8) + # sets thickness of line
  geom_boxplot( # Draw boxplots for each day
    aes(fill = day_of_week), # colors boxes by weekday
    width = 0.65, # sets width of boxes
    alpha = 0.45, # sets transparency level of boxes
    outlier.shape = NA) + # suppress default outlier points
  geom_jitter( # Add jittered points to show individual observations
    aes(color = day_of_week), # colors points by weekday
    width = 0.12, # sets horizontal jitter
    height = 0, # no vertical jitter
    alpha = 0.6, # sets transparency level
    size = 2) + # sets point size
  scale_fill_manual(values = week_colors, # Apply manual fill color scale
                    guide = 'none') + # removes legend
  scale_color_manual(values = week_colors, # Apply manual fill color scale
                     guide = "none") + # remove duplicate legend
  labs(
    title = "Sleep Duration by Day of Week", # sets plot title
    x = "Day of week", # sets x-axis title
    y = "Time Slept (minutes)") + # sets y-axis title
  base_theme + # uses preset base theme
  # Add curved arrow annotation highlighting highest median day
  geom_curve(
    data = NULL, # uses plot-level data
    aes(x = highlight_day , # sets arrow starting x-position 
        # sets arrow starting y-position 
        y = max(cleaned_data$time_slept_minutes, na.rm = TRUE) * 1.01, 
        xend = highlight_day , # sets arrow ending x-position 
        # sets arrow ending y-position 
        yend = med_by_day$med_sleep[med_by_day$day_of_week == highlight_day] * 1.05),
    curvature = -0.35, # sets degree of curve
    arrow = arrow(length = unit(0.1, "inches")), # sets arrowhead size
    inherit.aes = FALSE) + # prevents inheriting global aesthetics
  annotate( "text", # Add text label describing highlighted day
            x = highlight_day, 
            # sets vertical position of label
            y = max(cleaned_data$time_slept_minutes, na.rm = TRUE) * 1.03,
            # creates content of label
            label = paste0("Highest median: ", round(highlight_median, 0), " min"), 
            vjust = 0, # sets vertical alignment
            size = 3.4, # sets text size of label
            color = "gray20", # colors text
            family = "space") # sets font family of label

p1_box # shows plot 1

# saves plot 1 with specifications
ggsave("plot1_sleep_by_day.png", p1_box, width = 9, height = 5, dpi = 300)
```

# Plot 2: Scatter + Bins (Sleep duration vs efficiency)
```{r}
p2_sleep_scatter <- 
  ggplot(cleaned_data, # uses cleaned data
         aes(x = sleep_efficiency, # x-axis is continuous predictor (weekday)
             y = time_slept_minutes)) + # y-axis is response (sleep duration)
  geom_bin2d(bins = 10) +   # Create rectangular bins instead of individual points 
  scale_fill_gradientn(
    colours = palette_gradient,
    name = "Number of days")+ # creates legend title 
  labs(
    title = "Sleep Duration vs Efficiency", # sets plot title
    x = "Sleep Efficiency (0-1)", # sets x-axis title
    y = "Time slept (minutes)") + # sets y-axis title
  base_theme # uses preset base theme

p2_sleep_scatter # shows plot 2
# saves plot 2 with specifications
ggsave("plot2_duration_vs_efficiency.png", p2_sleep_scatter, width = 7.5, height = 5, dpi = 300)
```

# Plot 3: Calendar-style heatmap (Daily sleep over time)
```{r}
# Create variables needed for a calendar grid layout
cleaned_data_cal <- cleaned_data |>
  # set order Monday–Sunday so heatmap rows display chronologically
  mutate(weekday = factor(wday(date, label = TRUE, abbr = FALSE), 
                          levels = c("Monday",
                                    "Tuesday", 
                                    "Wednesday",
                                    "Thursday",
                                    "Friday",
                                    "Saturday",
                                    "Sunday")),,
    # Create a week index variable
    week = floor_date(date, unit = "week", week_start = 1))



# Identify the unusually low-sleep Monday
low_sleep_day <- cleaned_data_cal |> # uses altered calender plot data
  filter(time_slept_minutes == min(time_slept_minutes)) # selects outlier



# Create a variable flagging that extreme low day
cleaned_data_cal <- cleaned_data_cal |> # uses altered calender plot data
  mutate(low_sleep_flag = if_else(
      time_slept_minutes == min(time_slept_minutes),  # detects lowest sleep day
      "Extreme low sleep","Normal range")) # identifies sleep range as normal or low

# Compute gradient limits excluding the extreme day
sleep_range <- cleaned_data_cal |> # uses altered calender plot data
  filter(low_sleep_flag=="Normal range") |> # filters to 'normal' sleep durations
  summarise(
    min_sleep = min(time_slept_minutes), # sets min bound of range
    max_sleep = max(time_slept_minutes)) # sets max bound of range

p3_calendar <- 
  ggplot(cleaned_data_cal, # uses altered calender plot data
         aes(x = week, # column position = week grouping
             y = weekday)) + # row position = day of week
  geom_tile( # Draw rectangular tiles for each date (in NORMAL range)
    data = cleaned_data_cal # uses cleaned data
      |> filter(low_sleep_flag == "Normal range"), # filter to normal sleep range
    aes(fill = time_slept_minutes), # fills color by time slept
    color = "white", # makes white borders between tiles
    width = 6.5, # sets tile width
    height = 0.95, # sets tile height
    linewidth = 0.6) + # sets thickness of tile borders
  geom_tile( # Draw rectangular tiles for each date (in LOW range)
    data = cleaned_data_cal # uses cleaned data
      |> filter(low_sleep_flag == "Extreme low sleep"), # filters to outliers
    fill = 'green3', # colors outliers green
    color = "white", # makes white borders between tiles
    width = 6.5, # sets tile width
    height = 0.95, # sets tile height
    linewidth = 0.6) + # sets thickness of tile borders
  scale_x_date(date_labels = "%b %d") + # display axis format like "Jan 09"
  scale_fill_gradientn(
    colors = palette_gradientA, # Define color gradient for sleep duration
    name = "Sleep Duration (min)", # sets legend title
    limits = c(sleep_range$min_sleep, # sets upper limit for scale
               sleep_range$max_sleep)) + # sets lower limit for scale
  # flips axis so Monday appears at top of heatmap
  scale_y_discrete(limits = rev(levels(cleaned_data_cal$weekday))) +
  labs(
    title = "Sleep Calendar", # sets title
    caption = "Green tile = outlier low-sleep day (excluded from color-scale)",
    x = "Week (starting Monday)", # sets x-axis title
    y = "") + # no title for y-axis
  base_theme + # uses preset base theme
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1), # rotate week labels 
    panel.grid = element_blank()) # Remove background grid lines

p3_calendar # shows plot 3
# saves plot 3 with specifications
ggsave("plot3_calendar_heatmap.png", p3_calendar, width = 10, height = 4.8, dpi = 300)
```

# Plot 4: Bubble scatter (Steps vs sleep duration; color = time stressed)

```{r}
# Identify sleep outlier value (low sleep day)
sleep_outlier <- min(cleaned_data$time_slept_minutes, na.rm = TRUE)

# Flag outlier day
cleaned_data_bubble <- cleaned_data |> # uses cleaned data
  mutate(sleep_outlier_flag = time_slept_minutes ==sleep_outlier) # flags outlier

# Compute color scale limits excluding the outlier day
stress_limits <- cleaned_data_bubble |> # uses altered data for bubble plot
  filter(!sleep_outlier_flag) |> # filters to non-outlier data
  summarise(
    min_stress = min(time_stressed_min, na.rm = TRUE), # sets min border
    max_stress = max(time_stressed_min, na.rm = TRUE)) # sets max border

p4_bubble <-
  ggplot(cleaned_data_bubble, # uses altered data for bubble plot
         aes(x = number_of_steps, # x-axis is # of steps
             y = time_slept_minutes)) + # y-axis is response (sleep duration)
  geom_point( # creates non-outlier points
    data = cleaned_data_bubble # uses altered data for bubble plot
      |> filter(!sleep_outlier_flag), # selects only non-outliers
    aes(color = time_stressed_min), # colors by time stressed
    size = 3) + # sets size to 3
  geom_point( # creates outlier point excluded from color-scale
    data = cleaned_data_bubble # uses altered data for bubble plot
      |> filter(sleep_outlier_flag), # selects only outlier
    color = "green3", # colors outlier blue
    size = 3) + # sets size to 3
  scale_x_continuous(labels = comma) + # formats x-axis with commas
  scale_color_gradientn(
    colors = palette_gradientA, # custom gradient colors
    name = "Time stressed (min)", # legend title
    limits = c(stress_limits$min_stress, # sets min limit 
               stress_limits$max_stress)) + # sets max limit 
  labs(title = "Steps vs Sleep (Colored by Time Stressed)", # sets title
    caption = "Green point = low-sleep outlier", # creates caption
    x = "Steps", # sets x-axis title
    y = "Time stressed (minutes)") + # sets y-axis title
  base_theme # uses base theme

p4_bubble # shows plot 4
# saves plot 4 with specifications
ggsave("plot4_steps_stress_sleep.png", p4_bubble, width = 7.5, height = 5, dpi = 300)
```

# Plot 5: Lollipop chart (Mean sleep by hours worked)
```{r}
# Summary by hours worked 
cleaned_data_work_summary <- cleaned_data |> # uses cleaned data
  group_by(hours_worked) |> # groups observations by hours worked
  summarise(mean_sleep = mean(time_slept_minutes), # compute mean sleep per group
    n_days = n(), # counts number of days in each group
    .groups = "drop") # ungroups df

overall_mean_sleep <- mean(cleaned_data$time_slept_minutes) # finds overall mean

highlight_work <- cleaned_data_work_summary |> # uses summary by hours worked
  slice_max(mean_sleep, n = 1)  # selects row with highest mean_sleep

p5_lollipop <- 
  ggplot(cleaned_data_work_summary, # uses summary by hours worked
         aes(x = factor(hours_worked), # converts to factor for categorical x-axis
             y = mean_sleep)) + # y-axis shows group mean sleep
  geom_hline( # add dashed horizontal line showing overall mean sleep
    yintercept = overall_mean_sleep, # sets line at overall mean
    linetype = "dashed", # sets line type to dashed
    linewidth = 0.8) + # sets line width
  geom_segment( # adds vertical lollipop stems from 400 up to group mean
    aes(xend = factor(hours_worked), # keeps same x-position as point
        y = 0, # line starts at 0 min
        yend = mean_sleep), # line ends at mean for the group
    linewidth = 1.1) + # sets line width
  geom_point(size = 4) + # sets size of points (at mean for group)
  geom_text(
    aes(label = paste0("n=", n_days)),  # labels each point with # obs in group
    vjust = -1.2, # positions label
    size = 3.2, # sets size of label
    family = "space")+ # sets font family
  geom_label(data = highlight_work, #adds label highlighting highest mean
    aes(x = as.numeric(factor(hours_worked)), # sets x-axis position
      y = mean_sleep + 15, # sets y-axis position
      # writes label text
      label = paste0("Highest average sleep\n ( worked ", hours_worked, " hrs)")), 
    size = 3.2, # sets text size
    family = 'space',
    label.size = 0.2, # sets border size
    inherit.aes = FALSE) + # prevents inheriting global aesthetics
  labs(
    title = "Average Sleep by Hours Worked", # sets title
    subtitle = "Group means; labels show number of days per group", # sets subtitle
    x = "Hours worked", # sets x-axis title
    y = "Mean time slept (minutes)") + # sets y-axis title
  base_theme # uses base theme

p5_lollipop # shows plot 5
# saves plot 5 with specifications
ggsave("plot5_lollipop_hours_worked.png", p5_lollipop, width = 7.5, height = 5, dpi = 300)
```

